# Reproducible Research: Peer Assessment 1

```{r setoptions}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r loading_data}
rm(list = ls()) ##clear the environment

activity = read.csv("./activity.csv", header=TRUE, 
                    colClasses = c("numeric", "Date", "numeric"))
```


## What is mean total number of steps taken per day?
```{r stepsPerDay}
library(dplyr)

stepsPerDay = activity %>%
            group_by(date) %>%
            summarise( totalSteps = sum(steps, na.rm=TRUE))
paste("Mean # of Steps: ", round( mean(stepsPerDay$totalSteps), 1), " Median: ", round(median(stepsPerDay$totalSteps)));

```

```{r Histogram_stepsPerDay}
library(ggplot2)
p = ggplot(stepsPerDay, aes(x=totalSteps))
p + geom_histogram(aes(y = ..count..), binwidth=2000) +
    ggtitle("Steps Histogram") + 
    xlab("Total Steps")
```

## What is the average daily activity pattern?

```{r dailyIntervalActivity}

dailyPattern = activity %>%
    filter(! is.na(steps) ) %>%
    group_by(interval) %>%
    summarise(meanSteps = mean(steps) )


p = ggplot(dailyPattern, aes(interval, meanSteps))
p + geom_line() + 
    ggtitle("Daily Interval Activity") +
    ylab("Average Steps")

```

Maximum number of Steps:

```{r getInterval}
maxStep = dailyPattern %>%
            arrange(desc(meanSteps), interval) %>%
            head(1)
maxStep

```

The interval is `r maxStep$interval`
## Imputing missing values

```{r}
sum(is.na(activity$steps))
```

Normalizing the missing data with Average Values
```{r }
library(data.table)
activity = as.data.table(activity)
#filling the missing steps using the mean of steps for that interval
normalizedMissingData = activity[, steps:=ifelse(is.na(.SD$steps), 
                                               mean(.SD$steps, na.rm=TRUE), 
                                               .SD$steps), 
                                interval]
```

```{r Histogram_NormalizedData}
stepsPerDayNormalized = normalizedMissingData %>%
    group_by(date) %>%
    summarise(totalSteps  = sum(steps))

p = ggplot(stepsPerDayNormalized, aes(x=totalSteps))
p+geom_histogram(aes(y=..count..), binwidth = 2000) + 
    ggtitle("Total Steps Histogram") + 
    xlab("Total Steps")

paste("Mean # of Steps: ", round( mean(stepsPerDayNormalized$totalSteps), 1), 
      " Median: ", round( median(stepsPerDayNormalized$totalSteps) ))
```

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday_and_weekend_pattern}
# set the locale
#Sys.setlocale("LC_TIME", "C")

#Categorize Days of Week and Plot
wkdays = weekdays(normalizedMissingData$date, abbreviate=T)
type = ifelse(wkdays %in% c("Sat", "Sun"), "Weekend", "Weekday" )

weektype = normalizedMissingData %>% 
    mutate(type = factor(type) ) %>%
    group_by(interval, type) %>%
    summarise(totalSteps = sum(steps))

p = ggplot(weektype, aes(interval, totalSteps))
p + geom_line() + facet_wrap(~type, ncol=1) + 
    ggtitle("Weekend vs Weekday Activity") + 
    ylab("Average Steps")
```
